Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-09-23 grove-paper.tex
Wrote Section 2.3.6 Multi-User Cycles (Michael D. Adams)

While a single-user edits can always be checked to see
whether they create a cycle and thus the system
can immediately warn the user,
multi-user edits may not manifest cycles until after
the edits are merged.

For example, consider the situation shown in \autoref{fig:Multi-User Cycles}.
The top graph represents code that is synchronized between Alice and Bob
and represents the code~\texttt{(1 * \hole) * (2 * \hole)}.
In the left graph, Alice moves vertex~TODO to the \texttt{right}
index of vertex~TODO.
The resulting graph represents the code~\texttt{(1 * (2 * \hole)) * \hole}.
In the right graph, Bob moves vertex~TODO to the \texttt{right} index of vertex~TODO.
The resulting graph represents the code~\texttt{\hole * (2 * (1 * \hole))}.
On their own, neither edit creates a cycle.
However, when these edits are merged,
not only is a cycle created, but
the vertexes in the cycle

Can also be detected due to neither original leaving them disconnected.

\begin{figure*}
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
    "$\texttt{+}_{4}$" [> "left", > "19"'],
    "$\texttt{*}_{6}$" [> "left", > "19"']
   },
   {}
  }
 }
};
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
   },
   {
    "$\texttt{+}_{4}$" [> "left", > "19"'] -> {
     "$\texttt{*}_{6}$" [> "left", > "19"']
    }
   }
  }
 }
};
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-b}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
   },
   {
    "$\texttt{*}_{6}$" [> "left", > "19"'] -> {
     "$\texttt{+}_{4}$" [> "left", > "19"']
    }
   }
  }
 }
};
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-c}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
   },
   {
    v1/"$\texttt{*}_{6}$" [> "left", > "19"'],
    v2/"$\texttt{+}_{4}$" [> "left", > "19"']
   }
  }
 }
};
\path[draw] (v1) -> (v2);
\path[draw] (v2) -> (v1);
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-d}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Multi-User Cycles}
\label{fig:Multi-User Cycles}
\end{figure*}

@@

Each edge, $\e = (u, v_s, p, v_t)$, has a unique name, $u \in \U$, and connects the child position $p$ of a source vertex $v_s$ to a target vertex $v_t$.
The state of each edge, $\e$, is determined by the corresponding edge state, $\G(\e) \in \Sigma = \{ \bot, \Plus, \Minus \}$. 
If $\G(\e) = \bot$, then $\e$ does not yet exist.
If $\G(\e)$ is $\mathbf{+}$, the edge has been created.
If $\G(\e)$ is -, the edge has been destroyed.
The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$.

@@

>>>>

While single-user edits can be checked to see
whether they create a cycle and thus the system
can immediately warn the user,
multi-user edits may not manifest cycles until after
the edits from multiple users are merged.

For example, consider the situation in \autoref{fig:Multi-User Cycles}.
\autoref{fig:Multi-User Cycles-a} shows the graph after Alice and Bob have made some more edits,
and Alice and Bob's editors are synchronized with each other.
This graph represents the expression~\texttt{((\hole+\hole)+(\hole*\hole))+\hole}.
Alice moves the leftmost \texttt{+} into the rightmost hold
and then the \texttt{*} underneath that.
This results in \autoref{fig:Multi-User Cycles-b},
which represents the expression~\texttt{(\hole+\hole)+((\hole*\hole)+\hole)}.
On the other hand, Bob does the same move
but puts the \texttt{+} and under the \texttt{*}.
This results in \autoref{fig:Multi-User Cycles-c},
which represents the expression~\texttt{(\hole+\hole)+((\hole+\hole)*\hole)}.

On their own, neither of these edits creates a cycle.
However, merging the edit actions of both Alice and Bob results in the graph
in \autoref{fig:Multi-User Cycles-d}, which has a cycle
between \Vertex{TODO} and \Vertex{TODO}.
This situation is akin to a merge conflict in traditional version control systems.
Resolving it requires user input, but is as simple as
either deleting \Edge{TODO} and \Edge{TODO} (thus favoring Alice's version)
or deleting \Edge{TODO} and \Edge{TODO} (thus favoring Bob's version).
Since a cycle like this that is connected to the root vertex always
contains vertexes with multiple parents that if removed would break the cycle,
we use the user interface described in \autoref{sub:Multi-parent conflicts}
to display these kinds of programs.
For example, \autoref{fig:Multi-User Cycles-d} would be displayed
as shown in TODO:FIGURE.
We leave the user interface considerations of resolving the conflict such a
cycle represents to future work.

\begin{figure*}
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
    "$\texttt{+}_{4}$" [> "left", > "19"'],
    "$\texttt{*}_{6}$" [> "right", > "19"']
   },
   {}
  }
 }
};
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-a}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
   },
   {
    "$\texttt{+}_{4}$" [> "left", > "19"'] -> {
     "$\texttt{*}_{6}$" [> "left", > "19"'],
     {}
    }
   }
  }
 }
};
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-b}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
   },
   {
    "$\texttt{*}_{6}$" [> "left", > "19"'] -> {
     "$\texttt{+}_{4}$" [> "left", > "19"'],
     {}
    }
   }
  }
 }
};
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-c}
\end{subfigure}
%%%%%%%%
\hfill
%%%%%%%%
\begin{subfigure}{0.24\textwidth}
\centering
\begin{tikzpicture}
\path (0cm,0cm) graph[graph style] {
 root[root style] -> {
  "$\texttt{+}_{8}$" [> "root", > "7"'] -> {
   "$\texttt{+}_{2}$" [> "left", > "19"'] -> {
   },
   {
    v1/"$\texttt{*}_{6}$" [> "left", > "19"'],
    v2/"$\texttt{+}_{4}$" [> "left", > "19"']
   }
  }
 }
};
\path[draw,->] (v1) to [out=-45,in=-135] (v2);
\path[draw,->] (v2) to [out=-90,in=-90] (v1);
\end{tikzpicture}
\caption{Initial code}
\label{fig:Multi-User Cycles-d}
\end{subfigure}
%%%%%%%%
\hfill{}
\caption{Multi-User Cycles}
\label{fig:Multi-User Cycles}
\end{figure*}

@@

Each edge, $\e = (u, v_s, p, v_t)$, has a unique name, $u \in \U$, and connects the child position $p$ of a source vertex $v_s$ to a target vertex $v_t$.
The state of each edge, $\e$, is determined by the corresponding edge state, $\G(\e) \in \Sigma = \{ \bot, \Plus, \Minus \}$. 
If $\G(\e) = \bot$, then $\e$ does not yet exist.
If $\G(\e)$ is $\Plus$, the edge has been created.
If $\G(\e)$ is $\Minus$, the edge has been destroyed.
The total ordering $\bot \sqsubset \Plus \sqsubset \Minus$ forms a lattice over $\Sigma$.

@@

\section{Supplemental Material: Complete Graph Sequences for All Figures}
\label{apx:Supplemental Material: Complete Graph Sequences for All Figures}

TODO: put all intermediate states for all graphs along with the edge actions
